---
title: "Challenge 8 Solution - SNL Gender Distribution Over Time"
author: "Linus Jen"
description: "Joining Data"
date: "6/27/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - snl
  - Linus Jen
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(here)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Challenge Overview

For this challenge, I will be investigating the SNL related datasets (`snl_actors.csv`, `snl_casts.csv`, and `snl_seasons.csv`).

For my analysis, I want to look at how the gender diversity has changed over time.

```{r}
actors <- read_csv(here("posts", "_data", "snl_actors.csv"))
casts <- read_csv(here("posts", "_data", "snl_casts.csv"))
seasons <- read_csv(here("posts", "_data", "snl_seasons.csv"))

# Rename some columns
casts <- casts %>%
  rename("n_eps_present" = "n_episodes",
         "first_ep_present" = "first_epid",
         "last_ep_present" = "last_epid")
seasons <- seasons %>%
  rename("s_first_epid" = "first_epid",
         "s_last_epid" = "last_epid",
         "s_n_episodes" = "n_episodes")

# Summary information
print(summarytools::dfSummary(actors,
                              varnumbers=FALSE,
                              plain.ascii=FALSE,
                              style="grid",
                              graph.magnif = 0.70,
                              valid.col=FALSE),
      method="render",
      table.classes="table-condensed")

print(summarytools::dfSummary(casts,
                              varnumbers=FALSE,
                              plain.ascii=FALSE,
                              style="grid",
                              graph.magnif = 0.70,
                              valid.col=FALSE),
      method="render",
      table.classes="table-condensed")

print(summarytools::dfSummary(seasons,
                              varnumbers=FALSE,
                              plain.ascii=FALSE,
                              style="grid",
                              graph.magnif = 0.70,
                              valid.col=FALSE),
      method="render",
      table.classes="table-condensed")
```

### Briefly describe the data

This data contains information about the casts and actors that partook in any SNL show.

`snl_actors.csv` contains 2,306 rows of people who were part of SNL, and holds the name (under `aid`), the `url` for the person, the `type` of member (cast, guest, crew, or unknown, and `gender` (male, female, unknown, and andy (not sure what gender andy is)). This data is already tidy.

`snl_casts.csv` contains information about cast members, and is broken down by the the name (under `aid`) and the season (`sid`). It also tells if the member was `featured` in the show or season, the first episode and last episode of SNL they were on (`first_epid` and `last_epid`, respectively), if they were part of the weekend `update_anchor`, the `n_epsiodes` they were on, and what fraction of episodes for a particular season they were featured on (`season_fraction`). Both `first_epid` and `last_epid` seem to be dates in the form of `YYYYMMDD`, but because this seems to be an identifier, I won't be changing it to a date type. This data also seems to already be tidy.

Lastly, `snl_seasons.csv` gives insight about each season, and there 46 years in this dataset (from 1975 to 2020). It includes a unique identifier (`sid`), `year`, IDs for the first and last episode (`first_epid` and `last_epid`), and `n_episodes` per season. both `first_epid` and `last_epid` seem to be dates in the form of `YYYYMMDD`, but because this seems to be an identifier, I won't be changing it to a date type.

To join our data, I will be using the `aid` columns from `actors` and `casts` as the join condition, and then use the `sid` from `casts` and `seasons` to join these two datasets. I will use an inner join, as for my analysis comparing gender diversity across the history of SNL, I need gender information (found in `actors`) and year that each actor was present (found in `seasons`)


## Joining and tidying the data

Given that `casts` is the dataframe connecting all datasets together, we expect to see each row representing a cast member and a season. Thus, there should be at most 614 rows (matching the rows of `casts`), though it can be less if there aren't any joins. We also expect 15 columns (4 columns in `actors`, 8 columns in `casts`, and `5` columns in `seasons`, minus 2 ID columns from joining).

```{r}
# Join datasets
full_df <- casts %>%
  inner_join(actors, by="aid") %>%
  inner_join(seasons, by="sid")

# View data together
print(summarytools::dfSummary(full_df,
                              varnumbers=FALSE,
                              plain.ascii=FALSE,
                              style="grid",
                              graph.magnif = 0.70,
                              valid.col=FALSE),
      method="render",
      table.classes="table-condensed")
```

From the summary statistics above, we see that there are 614 columns rows and 15 columns, which matches our predictions.

## Analysis

As mentioned previously, my goal is to see how the gender demographics change over time. To do this, I will make a line plot with the x-axis representing the year, and the y-axis representing the proportion of a specific gender who made up the cast. Each line will represent a `gender` category.

```{r}
gender_breakdown <- full_df %>%
  group_by(year, gender) %>%
  summarise(n_gender = n_distinct(aid)) %>%
  ungroup() %>%
  group_by(year) %>%
  mutate(prop_gender = n_gender / sum(n_gender))

# Show dataste
gender_breakdown

# Visualize data
ggplot(gender_breakdown, aes(x=year, y=prop_gender, color=gender)) +
  geom_line() +
  geom_point() +
  geom_text(aes(x=year, y=prop_gender, label=n_gender), vjust=-0.5) +
  theme_minimal() +
  scale_x_continuous(breaks=min(gender_breakdown$year):max(gender_breakdown$year), 
                     labels=min(gender_breakdown$year):max(gender_breakdown$year)) +
  scale_y_continuous(labels=scales::percent_format(accuracy=1)) +
  labs(title="SNL Gender Distribution Across the Year",
       x="Year",
       y="Percentage (%)") +
  theme(axis.text.x = element_text(angle=90)) +
  guides(fill=guide_legend(title="Gender"))
```

The graph above shows both the percentages and the counts the genders for the casts for each show, assuming that the data is correct. It's interesting to see that the proportion of males on the show tends to decrease over time, while the proportion of women tends to increase over time. The number of cast members also seems to increase over time, as in 1975, there were only 9 distinct people in the entire season, but by 2020, there were 20 unique people on the show. 
